Unbalanced Mallows Models for Optimizing Expensive Black-Box Permutation Problems

### 2021-02-05 paper/main.tex
final updates

\usepackage{booktabs} % For formal tables
\usepackage{enumitem}
\usepackage{afterpage,tabularx}
\hypersetup{draft}
%\usepackage[colorlinks=true, allcolors=blue]{hyperref}

@@

We consider three different values $\FEmax \in\{100, 200,400\}$ for the maximum budget of  evaluations of the
actual objective function. In a white-box context, state-of-the-art algorithms
for the LOP and PFSP typically evaluate thousands of solutions,
thus, the budget considered here for the black-box context is extremely
limited.

@@

CEGO replaces the Euclidean distance measure, used by the surrogate model in
EGO, with a distance measure appropriate to combinatorial
landscapes~\citep{ZaeStoBar2014:ppsn}.
% , such as Kendall's-$\tau$ distance for permutations
% (Eq.~\ref{eq:kendall_decomp}).
In CEGO, the surrogate model is
explored by a GA with crossover and mutation operators appropriate for the
particular combinatorial problem. The original paper notes that coupling the GA
with local search does not improve the results significantly since the model is
anyway an inexact estimation of the original fitness
function~\citep[p.~875]{ZaeStoFriFisNauBar2014}. A thorough comparison of CEGO
with other optimizers for expensive black-box combinatorial optimization ranked
it as the best-performing~\citep{ZaeStoFriFisNauBar2014}.

We use the original implementation of GECO.\footnote{\url{https://cran.r-project.org/package=CEGO}} %
Although it is never stated in the original paper, the implementation of CEGO
generates a set of initial solutions of size $\minit=10$ by means of a
max-min-distance sequential design: new solutions are added to the set
sequentially by maximizing the minimum distance to solutions already in the
set. These initial solutions are then evaluated on the true fitness function
and the result is used to build the initial surrogate
model. % \citet{ZaeStoBar2014:ppsn} uses \minit=10
Following the authors of
CEGO~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn}, we use a GA to optimize
the surrogate models with population size of 20, crossover rate of 0.5,
mutation rate of $1/n$, tournament selection of size 2 with probability of 0.9,
interchange mutation (i.e., exchanging two randomly selected elements) and
cycle crossover for permutations. To generate each new solution, CEGO runs the
GA with a budget of $10^4$ evaluations of the model as a surrogate of the true fitness function.  
%and we use the same configuration parameters.% However, with such a setting, a single run of CEGO with a budget of
%$\FEmax=400$ true fitness evaluations requires \emph{more than a week} of
%computation time, which was infeasible for our study. 
%Due to extremely long
%runtimes of CEGO and constraints in our computing system, we set the budget of
%the GA to $10^3$ and we stopped each run of CEGO after 144 hours of CPU-time.
%Although a higher GA budget could improve the results of CEGO reported here,
%our results are similar (and sometimes better) than the ones reported in the
%original papers~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn} due to
%running CEGO with $\FEmax=400$ instead of $\FEmax=200$. We found that
%doubling $\FEmax$ produces a smaller increase in runtime than multiplying the
%GA budget by a factor of $10$. Nevertheless, we caution that choosing a
%different trade-off may lead to different conclusions and more work is needed
%to understand CEGO parameters.

\myparagraph{Other settings}
%
For simplicity, we use Kendall's-$\tau$
distance~(Eq.~\ref{eq:kendall_decomp}) in all experiments.
%Nevertheless,
%\citet{ZaeStoFriFisNauBar2014} points out that CEGO performs best on the QAP
%when using Hamming distance and the same may apply to UMM.
We plan to extend UMM to other distance measures in the future.

@@

@@

From this preliminary experiment, it is not obvious how the best settings are
related to problem features, such as permutation length ($n$). On the other
hand, given the sharp transitions in the plots, it is clear that further
fine-tuning may improve performance and uncover further patterns.

@@

\subsection{Rankings and orderings: experiments}\label{sec:exper_ro}

@@

%During our experiments, we realized that the \emph{swap} distance used by CEGO
%assumes rankings, i.e., it actually measures the \ken distance, yet CEGO did
% not invert the permutation before evaluating the PFSP.
During our experiments, we realized that CEGO claimed to use the Swap distance. Instead, CEGO implements the \ken distance, yet CEGO did
not invert the permutation before evaluating the PFSP.
%\MANUEL{no me convence esta forma de expresarlo porque la crítica obvia sería que por qué invertimos la función objectivo y no la función de distancia?} \EKHINE{es lo mismo}
We verified that this is
indeed how the implementation works and that those results match the published
ones~\citep{ZaeStoBar2014:ppsn}. 
Note that using the Swap distance as claimed by CEGO is equivalent to using \ken distance and inverting the permutation before calling the objective
function. Thus, the obvious question arises about the effect of inverting the permutation before calling the objective
function, so that the rankings are transformed to ordering, as done by our UMM
algorithm. In the following, we show results of both variants, that is, the original CEGO (\CEGOorig) and our modified version that inverts the permutation before evaluation (\CEGOinv).

@@

@@

Figure~\ref{fig:results} shows the results on the real-world instances from LOLIB and PFSP. There is one figure per instance. In each figure there is one horizontal line displaying the best known solutions for the problem. Besides, there are 5 different lines representing each of the different algorithm/versions considered, in particular:
\begin{itemize}
\item UMM$_{100}$, UMM$_{200}$ and UMM$_{400}$ are three different versions for UMM, each considering a different total number of function evaluations, 100, 200 and 400. Obviously, the lines corresponding to UMM$_{100}$, UMM$_{200}$ wont reach the right corner of the figure. As mentioned in Sec.~\ref{sec:sampling}, the behavior of
UMM depends on the value of $m$. 
\item \CEGOorig and \CEGOinv are, respectively, the original version of CEGO and the modification of Section~\ref{sec:exper_ro}, i.e., the ranking generated by the CEGO algorithm at each iteration, $\sigma$, is inverted before being evaluated, so the evaluation is performed on $\sigma^{-1}$. 
\end{itemize}

Since there is a small number of evaluations and the algorithms evaluate just
one solution at each step, we chose to record and plot the fitness of the
solution evaluated at each step of each run, instead of the usual plots of the
best-so-far solution over number of evaluations. We believe that plotting each
evaluated solution provides better insights on the search dynamics of the
algorithms analyzed here, despite it is not directly comparable with most literature results. Please note that the algorithms do not return the
final solution shown in the plots, but rather the best one (minimum fitness) of
each run. Nevertheless, the minimum values achieved in the plot up to a
particular number of function evaluations gives an estimation of the mean
fitness of the best-so-far solution up to that point in the run. We provide
later an analysis of the best solution found by each run of the algorithm, which are comparable with the original paper of CEGO.

Figure~\ref{fig:results} shows such plots for a few LOP instances from LOLIB
and the PFSP.  For each problem instance and each algorithm, we plot the mean
fitness of the solution evaluated at each step over 10 runs with different
random seeds. The horizontal line indicates the best-known fitness for that instance.
% 
The shaded area shows one standard deviation around the mean. Ideally, each new
evaluation will monotonically improve in fitness, since each solution evaluated
helps to refine the model and leads to a better solution being evaluated in the
next step.  However, it is possible that new solutions are worse than their
predecessors. %Nevertheless, in this example, there is a clear
% overall improvement as more evaluations are performed.

\myparagraph{Convergence analysis of UMM} We start our analysis with the three
configurations of UMM. We can see a common trend in all instances: After
the 10 random initial permutations, the fitness of new solutions decreases
with a linear trend along iterations. The (close to) linear decrease is given
by the linear decrease in the expected distance of the model, as described in Section~\ref{sec:sampling}. Overall, the rate of improvement
does not seem to converge, not even for the longer runs of 400 evaluations. Therefore, we conjecture that longer runs are likely to produce further improved solutions. Moreover, a non-linear decrease in the expected distance should result in a faster convergence for larger evaluation budgets.

\myparagraph{Convergence analysis of CEGO}
The red and the gold lines in the plots shown in Fig.~\ref{fig:results}
correspond to the two variants of CEGO the original \CEGOorig and our modified
version \CEGOinv. The different between these two variants is overwhelmingly in
favour of \CEGOinv. With the exception of the \texttt{reC19} PFSP instance, the
search behavior of \CEGOorig resembles a random search, after a brief
improvement phase at the start of the run. By contrast, \CEGOinv shows a very
fast improvement in fitness within the first 50 evaluations that tends to
converge around the 150$^\text{th}$ evaluation, showing little improvement
after that. Since the only difference between these two variants is the
transformation from rankings to orderings before evaluating the objective
function (Sec.~\ref{sec:exper_ro}), we must conclude that this difference has
tremendous effect in performance. Therefore, we will focus our analysis on
\CEGOinv.

\myparagraph{Comparison of UMM and CEGO}
When comparing the UMM and CEGO configurations shown in Fig.~\ref{fig:results},
we can observe that all UMM configurations perform much better than the
original CEGO (\CEGOorig), while they converge slower than our modified
\CEGOinv. However, the rate of improvement of the UMM variants does not seem to
slow down near the maximum budget given and, in several instances such as
\texttt{rec31} and \texttt{N-sgb75-01}, matches and surpasses the solutions
found by \CEGOinv. Interestingly, this is the case, even for the shorter budget
configurations of UMM, i.e., UMM$_{100}$ and UMM$_{200}$. We also observe a
pattern that UMM performs better, or \CEGOinv performs worse, on larger
instances, as can be seen by comparing \texttt{rec19} with \texttt{rec31}.

\myparagraph{Comparison of the best solutions found}
%
We now turn our analysis to the best (minimum) fitness found at the end of the
run by the two best algorithms, that is, \CEGOinv and UMM$_{400}$. We show in
Table~\ref{tab:results} the mean fitness (and standard deviation) of the best
solution found over the 10 runs of each algorithm, together with a 95\%
confidence interval (CI) around their mean fitness difference. 
Except for PFSP
instances \texttt{rec05} and \texttt{rec31}, the differences are statistically
significant in favour of \CEGOinv (the intervals do not contain the value
zero). Nevertheless, the CIs indicate that the differences, although
consistent, are relatively small.
The detailed results of UMM$_{100}$  and UMM$_{200}$  and the original \CEGOorig are given in Table~\ref{tab:results_fitness}. In this table, it can be seen that the quality of \CEGOorig  solutions is similar to the UMM$_{100}$ version of UMM.

\myparagraph{Runtime comparison}
%
The average
running time is approximately 22 hours for the 400 function evaluations, which
is 300 times slower than UMM for the same number of function evaluations.

Table~\ref{tab:results} shows a brief comparison of time and the full ones are given in Table~\ref{tab:results_time}. The difference is time in impressive. 
The average runtime for all the instances is around 0.3 minutes, 1.15 minutes and 4.5 minutes respectively for the  versions with 100, 200 and 400 fitness evaluations. 
For CEGO in both versions the average time for 400 evaluations is more than 22 hour. This is 300 times slower than UMM for the same number of function evaluations. 
Moreover, UMM scales good for larger instances in quality performance. 
The computational complexity of UMM is $O(n \log n)$ at each iteration which causes the abrupt difference in the time performance of the algorithms. The details can be seen in Table~\ref{tab:results_time}.

\begin{figure*}[tb]
\small
\centering%
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p40-01}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-02} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p40-02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{N-t59b11xx} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_IO_N-t59b11xx}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-t59d11xx} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_IO_N-t59d11xx}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{SGB\_N-sgb75\_01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_SGB_N-sgb75_01}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{SGB\_N-sgb75\_01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_SGB_N-sgb75_02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC05} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec05_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC13} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec13_txt}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC19} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec19_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC31} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec31_txt}
\end{minipage}
\label{fig:results}
\caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on various instances.\label{fig:results}}
\end{figure*}

@@

\newcommand{\mcolc}[2]{\multicolumn{#1}{c}{\bf #2}}
\begin{table*}[tb]
  \caption{Mean fitness (and standard deviation) of the best solution found at the end of each run, over 10 independent runs for each problem instance. The 95\% confidence interval corresponds to the two independent samples t-test on the mean difference between the fitness of \CEGOinv minus the one of UMM$_{400}$. 
  Mean runtime is shown in hours. 
  %CEGO runs were stopped after 144 hours. This was the case for all LOP instances.
  \label{tab:results}}
 % \resizebox{\textwidth}{!}{%
\begin{tabular}{r@{\hskip -1ex}r*{2}{r@{\hskip 1ex}l}r@{\hskip 0.5ex}lr@{\hskip 0.75ex}r}
 \toprule
            &              & \mcolc{4}{Mean fitness (sd)} & \mcolc{2}{\multirow[b]{2.45}{*}{\shortstack{95\% CI of the                                                                                       \\mean difference}}} & \mcolc{2}{Mean runtime} \\\cmidrule(lr){3-6}\cmidrule(lr){9-10}
\bf Problem & \bf Instance & \mcolc{2}{\CEGOinv}              & \mcolc{2}{UMM$_{400}$}                                                                  &           &           & \bf \CEGOinv                  & \bf UMM$_{400}$    \\\midrule
LOP         & N-p40-01     & 10496.6                      & (116.8)                                                                         & 12369.1   & (389.9)   & {[}-2157.1,& -1587.9{]}    & 19.9 & 0.3 \\
         & N-p40-02     & 11714.7                      & (84.7)                                                                          & 13160.5   & (454.8)   & {[}-1773.5,& -1118.1{]}    & 20.1 & 0.3 \\
         & N-t59b11xx   & 102408.0                     & (3800.2)                                                                        & 114075.5  & (5389.2)  & {[}-16084.2,& -7250.8{]}   & 21.1 & 0.4 \\
         & N-t59d11xx   & 12692.2                      & (1097.9)                                                                        & 16662.9   & (2155.3)  & {[}-5618.5,& -2322.9{]}    & 21.2 & 0.4 \\
         & N-sgb75.01   & 1893533.2                    & (28594.6)                                                                       & 1992059.9 & (40235.7) & {[}-131577.1,& -65476.3{]} & 33.2 & 1.0 \\
         & N-sgb75.02   & 1895801.8                    & (82847.9)                                                                       & 1967132.3 & (49228.1) & {[}-136421.2,& -6239.8{]}  & 33.3 & 1.1 \\
         \midrule
PFSP        & rec05        & 1276.2                       & (15.1)                                                                          & 1276.3    & (8.4)     & {[}-11.8,& 11.6{]}         & 18.3 & 0.1 \\
       & rec13        & 1992.9                       & (27.6)                                                                          & 2057.8    & (22.8)    & {[}-88.8,& -41.0{]}        & 18.5 & 0.1 \\
       & rec19        & 2264.5                       & (45.0)                                                                          & 2304.8    & (37.7)    & {[}-79.4,& -1.2{]}         & 18.2 & 0.2 \\
       & rec31        & 3415.5                       & (62.1)                                                                          & 3419.3    & (37.0)    & {[}-52.6,& 45.0{]}         & 23.1 & 0.5 \\
\bottomrule
\end{tabular}
%}
\end{table*}

% The situation is very similar for the QAP, where both algorithms show the same
% behaviour as in the PFSP for all instances tested, except for
% \texttt{kra} instances. As shown in Fig~\ref{fig:kra32}, for \texttt{kra32} both
% algorithms are able to improve over the number of evaluations, although not
% with ease. As in the synthetic LOP instances, we again observe that CEGO does
% improve faster than UMM initially, yet CEGO appears to get stuck around
% evaluation 250, and UMM overtakes it around evaluation 300. We believe the
% reason is the same as before, that is, the CEGO model is not able to adequately
% reflect the ruggedness of the actual landscape and, hence, the best solutions
% found by the GA searching the model are not improved solutions for the actual
% problem. Interestingly, there are well-known differences between
% \texttt{kra30a}, \texttt{kra30b}, \texttt{kra32}, on one hand, and \texttt{nug12}, \texttt{nug30} and
% \texttt{tho32}, on the other~\citep{AnsBriGou2002qap}. However, we cannot at
% this moment point out the root cause of this different behavior.

>>>>

\usepackage{booktabs} % For formal tables
\usepackage{enumitem}
\usepackage{afterpage,tabularx}
%\hypersetup{draft}
%\usepackage[colorlinks=true, allcolors=blue]{hyperref}

@@

@@

CEGO replaces the Euclidean distance measure, used by the surrogate model in
EGO, with a distance measure appropriate to combinatorial
landscapes~\citep{ZaeStoBar2014:ppsn}.
% , such as Kendall's-$\tau$ distance for permutations
% (Eq.~\ref{eq:kendall_decomp}).

In CEGO, the surrogate model is
explored by a GA with crossover and mutation operators appropriate for the
particular combinatorial problem. A thorough comparison of CEGO
with other optimizers for expensive black-box combinatorial optimization ranked
it as the best-performing~\citep{ZaeStoFriFisNauBar2014}.

We use here the original implementation of
GECO.\footnote{\url{https://cran.r-project.org/package=CEGO}} %
Although it is never stated in the original paper, the implementation of CEGO
generates a set of initial solutions of size $\minit=10$ by means of a
max-min-distance sequential design: new solutions are added to the set
sequentially by maximizing the minimum distance to solutions already in the
set. These initial solutions are then evaluated on the true fitness function
and the result is used to build the initial surrogate
model. % \citet{ZaeStoBar2014:ppsn} uses \minit=10
Following the authors of
CEGO~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn}, the surrogate model is optimized by a GA with  population size of 20, crossover rate of 0.5,
mutation rate of $1/n$, tournament selection of size 2 with probability of 0.9,
interchange mutation (i.e., exchanging two randomly selected elements) and
cycle crossover for permutations. To generate each new solution, CEGO runs the
GA with a budget of $10^4$ evaluations of the model as a surrogate of the true
fitness function.  The original paper~\citep[p.~875]{ZaeStoFriFisNauBar2014} notes that coupling the GA with local
search does not improve the results significantly since the model is anyway an
inexact estimation of the original fitness
function.

%and we use the same configuration parameters.% However, with such a setting, a single run of CEGO with a budget of
%$\FEmax=400$ true fitness evaluations requires \emph{more than a week} of
%computation time, which was infeasible for our study. 
%Due to extremely long
%runtimes of CEGO and constraints in our computing system, we set the budget of
%the GA to $10^3$ and we stopped each run of CEGO after 144 hours of CPU-time.
%Although a higher GA budget could improve the results of CEGO reported here,
%our results are similar (and sometimes better) than the ones reported in the
%original papers~\citep{ZaeStoFriFisNauBar2014,ZaeStoBar2014:ppsn} due to
%running CEGO with $\FEmax=400$ instead of $\FEmax=200$. We found that
%doubling $\FEmax$ produces a smaller increase in runtime than multiplying the
%GA budget by a factor of $10$. Nevertheless, we caution that choosing a
%different trade-off may lead to different conclusions and more work is needed
%to understand CEGO parameters.

@@

\myparagraph{Other settings}
%
We consider three different values $\FEmax \in\{100, 200,$ $400\}$ for the maximum
budget of evaluations of the actual objective function in UMM. In the case of
CEGO, we only run experiments with the largest budget $m=400$.  In a white-box
context, state-of-the-art algorithms for the LOP and PFSP typically evaluate
thousands of solutions, thus, the budget considered here for the black-box
context is extremely limited.
%
For simplicity, we use Kendall's-$\tau$
distance~(Eq.~\ref{eq:kendall_decomp}) in all experiments.
%Nevertheless,
%\citet{ZaeStoFriFisNauBar2014} points out that CEGO performs best on the QAP
%when using Hamming distance and the same may apply to UMM.
We plan to extend UMM to other distance measures in the future.

@@

From this preliminary experiment, it is not obvious how the best settings are
related to problem features, such as permutation length $n$. On the other
hand, given the sharp transitions in the plots, it is clear that further
fine-tuning may improve performance and uncover further patterns.

@@

\subsection{Experiments with rankings and orderings}\label{sec:exper_ro}

@@

% not invert the permutation before evaluating the PFSP.
% During our experiments, we realized that CEGO claimed to use the Swap distance. Instead, CEGO implements the \ken distance, yet CEGO did
During our experiments, we realized that distance metric called Swap in CEGO~\citep{ZaeStoBar2014:ppsn} actually assumes rankings, that is, it actually implements the \ken distance, yet CEGO does
not invert the permutation before evaluating the PFSP.
%\MANUEL{no me convence esta forma de expresarlo porque la crítica obvia sería que por qué invertimos la función objectivo y no la función de distancia?} \EKHINE{es lo mismo}
We verified that this is
indeed how the implementation works and that those results match the published
ones~\citep{ZaeStoBar2014:ppsn}. 
%Note that using the Swap distance as claimed by CEGO is equivalent to using \ken distance and inverting the permutation before calling the objective
%function.
Thus, the obvious question arises about the effect of inverting the permutation before calling the objective
function, so that the rankings are transformed to orderings, as done by our UMM
algorithm. In the following, we show results of both variants, that is, the original CEGO (\CEGOorig) and our modified version that inverts the permutation before evaluation (\CEGOinv).

@@

\afterpage{
\begin{figure*}
\small
\centering%
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p40-01}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-p40-02} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_RandB_N-p40-02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{N-t59b11xx} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_IO_N-t59b11xx}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{N-t59d11xx} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_IO_N-t59d11xx}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{SGB\_N-sgb75\_01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_SGB_N-sgb75_01}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{SGB\_N-sgb75\_01} (LOP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_lop_SGB_N-sgb75_02}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC05} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec05_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC13} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec13_txt}
\end{minipage}
\\
\begin{minipage}{0.49\linewidth}
  \texttt{reC19} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec19_txt}
\end{minipage}
\begin{minipage}{0.49\linewidth}
  \texttt{reC31} (PFSP)\\[-0.5ex]
  \includegraphics[width=\textwidth]{../img/fitness_pfsp_rec31_txt}
\end{minipage}
\label{fig:results}
\caption{Mean fitness  (and standard deviation)  of each solution evaluated by each algorithm on various instances.\label{fig:results}}
\vspace{10em}
\end{figure*}
}

@@

Since the number of evaluations is  small and the algorithms evaluate just
one solution at each step, we chose to record and plot the fitness of the
solution evaluated at each step of each run, instead of the usual plots of the
best-so-far solution over number of evaluations. We believe that plotting each
evaluated solution provides better insights on the search dynamics of the
algorithms analyzed here. %, despite it is not directly comparable with most literature results.
Please note that the algorithms do not return the
final solution shown in the plots, but rather the best one (minimum fitness) of
each run. Nevertheless, the minimum values achieved in the plot up to a
particular number of function evaluations gives an estimation of the mean
fitness of the best-so-far solution up to that point in the run. We provide
later an analysis of the best solution found by each run of the algorithms.
%, which are comparable with the original paper of CEGO.

Figure~\ref{fig:results} shows the results on the real-world instances from
LOLIB and PFSP.  There is one plot per instance.  The horizontal line indicates
the best-known fitness for that instance.  Besides, there are 5 different lines
representing each of the different algorithm variants tested, in particular:
\begin{itemize}
\item UMM$_{100}$, UMM$_{200}$ and UMM$_{400}$ are three different versions of UMM, each considering a different total budget of function evaluations. As can be expected, the lines corresponding to UMM$_{100}$ and UMM$_{200}$ do not reach the right corner of the figure. As mentioned in Sec.~\ref{sec:sampling}, the behavior of
UMM depends on the value of $m$. 
\item \CEGOorig and \CEGOinv are, respectively, the original version of CEGO and the modification discussed in Section~\ref{sec:exper_ro}, i.e., the ranking generated by the CEGO algorithm at each iteration, $\sigma$, is inverted before being evaluated, so the evaluation is performed on $\sigma^{-1}$. 
\end{itemize}
For each algorithm, we plot the mean
fitness of the solution evaluated at each step over 10 runs with different
random seeds.
%Figure~\ref{fig:results} shows such plots for a few LOP instances from LOLIB
%and the PFSP.
% 
The shaded area shows one standard deviation around the mean. Ideally, each new
evaluation will monotonically improve in fitness, since each solution evaluated
helps to refine the model and leads to a better solution being evaluated in the
next step.  However, it is possible that new solutions are worse than their
predecessors. %Nevertheless, in this example, there is a clear
% overall improvement as more evaluations are performed.

\myparagraph{Convergence analysis of UMM} We start our analysis with the three
variants of UMM. We can see a common trend in all instances: After
the 10 random initial permutations, the fitness of new solutions decreases
with a linear trend along iterations. The (close to) linear decrease is given
by the linear decrease in the expected distance of the model, as described in Section~\ref{sec:sampling}. Overall, the rate of improvement
does not seem to converge, not even for the longer runs of 400 evaluations. Therefore, we conjecture that longer runs are likely to produce further improved solutions. Moreover, a non-linear decrease in the expected distance should result in a faster convergence for larger evaluation budgets.

\myparagraph{Convergence analysis of CEGO}
The red and the gold lines in the plots shown in Fig.~\ref{fig:results}
correspond to the two variants of CEGO the original \CEGOorig and our modified
version \CEGOinv. The difference between these two variants is overwhelmingly in
favour of \CEGOinv. With the exception of the \texttt{reC19} PFSP instance, the
search behavior of \CEGOorig resembles a random search, after a brief
improvement phase at the start of the run. By contrast, \CEGOinv shows a very
fast improvement in fitness within the first 50 evaluations and tends to
converge around the 150$^\text{th}$ evaluation, showing little improvement
after that. Since the only difference between these two variants is the
transformation from rankings to orderings before evaluating the objective
function (Sec.~\ref{sec:exper_ro}), we must conclude that this difference has
tremendous effect in performance.
%Therefore, we will focus our analysis on
%\CEGOinv.

\myparagraph{Comparison of UMM and CEGO}
When comparing the UMM and CEGO variants shown in Fig.~\ref{fig:results},
we can observe that all UMM variants perform much better than the
original CEGO (\CEGOorig), while they converge slower than our modified
\CEGOinv. However, the rate of improvement of the UMM variants does not seem to
slow down near the maximum budget given and, in several instances such as
\texttt{rec31} and \texttt{N-sgb75-01}, matches and surpasses the solutions
found by \CEGOinv. Interestingly, this is the case, even for the shorter budget
variants of UMM, i.e., UMM$_{100}$ and UMM$_{200}$. We also observe a
pattern that UMM performs better, or \CEGOinv performs worse, on larger
instances, as can be seen by comparing \texttt{rec19} with \texttt{rec31}.

\myparagraph{Comparison of the best solutions found}
%
We now turn our analysis to the best (minimum) fitness found at the end of the
run by the two best algorithms, that is, \CEGOinv and UMM$_{400}$. We show, in
Table~\ref{tab:results}, the mean fitness (and standard deviation), over the 10
runs of each algorithm, of the best solution found at the end of the run,
together with a 95\% confidence interval (CI) around their mean fitness
difference.  Except for PFSP instances \texttt{rec05} and \texttt{rec31}, the
differences are statistically significant in favour of \CEGOinv (the intervals
do not contain the value zero). Nevertheless, the CIs indicate that the
differences, although consistent, are relatively small.  Detailed results
of UMM$_{100}$ and UMM$_{200}$ and the original \CEGOorig are given in
Table~\ref{tab:results_fitness} for completeness. In this table, it can be seen that the quality
of \CEGOorig solutions is similar to the UMM$_{100}$ version of UMM.

\myparagraph{Runtime comparison}
%
Table~\ref{tab:results} also shows the runtime (in hours) of \CEGOinv and UMM when both
are run for 400 evaluations. For the same number of function evaluations, CEGO requires almost a day, whereas UMM
never requires more than one hour, often much less than that.
%
Details for other algorithm variants are given in Table~\ref{tab:results_time}.
The average runtime for all the instances is around 0.3 minutes, 1.15 minutes
and 4.5 minutes respectively for the UMM variants with 100, 200 and 400 fitness
evaluations.  For CEGO, in both versions, the average runtime for 400
evaluations is more than 22 hours. This is 300 times slower than UMM for the
same number of function evaluations.  Moreover, the runtime of UMM scales well
for larger instances since its computational complexity  is $O(n \log n)$ at
each iteration.

@@

\newcommand{\mcolc}[2]{\multicolumn{#1}{c}{\bf #2}}
\afterpage{%
\begin{table*}
  \caption{Mean fitness (and standard deviation) of the best solution found at the end of each run, over 10 independent runs for each problem instance. The 95\% confidence interval corresponds to the two independent samples t-test on the mean difference between the fitness of \CEGOinv minus the one of UMM$_{400}$. 
  Mean runtime is shown in hours. 
  %CEGO runs were stopped after 144 hours. This was the case for all LOP instances.
  \label{tab:results}}
 % \resizebox{\textwidth}{!}{%
\begin{tabular}{r@{\hskip -1ex}r*{2}{r@{\hskip 1ex}l}r@{\hskip 0.5ex}lr@{\hskip 0.75ex}r}
 \toprule
            &              & \mcolc{4}{Mean fitness (sd)} & \mcolc{2}{\multirow[b]{2.45}{*}{\shortstack{95\% CI of the                                                                                       \\mean difference}}} & \mcolc{2}{Mean runtime} \\\cmidrule(lr){3-6}\cmidrule(lr){9-10}
\bf Problem & \bf Instance & \mcolc{2}{\CEGOinv}              & \mcolc{2}{UMM$_{400}$}                                                                  &           &           & \bf \CEGOinv                  & \bf UMM$_{400}$    \\\midrule
LOP         & N-p40-01     & 10496.6                      & (116.8)                                                                         & 12369.1   & (389.9)   & {[}-2157.1,& -1587.9{]}    & 19.9 & 0.3 \\
         & N-p40-02     & 11714.7                      & (84.7)                                                                          & 13160.5   & (454.8)   & {[}-1773.5,& -1118.1{]}    & 20.1 & 0.3 \\
         & N-t59b11xx   & 102408.0                     & (3800.2)                                                                        & 114075.5  & (5389.2)  & {[}-16084.2,& -7250.8{]}   & 21.1 & 0.4 \\
         & N-t59d11xx   & 12692.2                      & (1097.9)                                                                        & 16662.9   & (2155.3)  & {[}-5618.5,& -2322.9{]}    & 21.2 & 0.4 \\
         & N-sgb75.01   & 1893533.2                    & (28594.6)                                                                       & 1992059.9 & (40235.7) & {[}-131577.1,& -65476.3{]} & 33.2 & 1.0 \\
         & N-sgb75.02   & 1895801.8                    & (82847.9)                                                                       & 1967132.3 & (49228.1) & {[}-136421.2,& -6239.8{]}  & 33.3 & 1.1 \\
         \midrule
PFSP        & rec05        & 1276.2                       & (15.1)                                                                          & 1276.3    & (8.4)     & {[}-11.8,& 11.6{]}         & 18.3 & 0.1 \\
       & rec13        & 1992.9                       & (27.6)                                                                          & 2057.8    & (22.8)    & {[}-88.8,& -41.0{]}        & 18.5 & 0.1 \\
       & rec19        & 2264.5                       & (45.0)                                                                          & 2304.8    & (37.7)    & {[}-79.4,& -1.2{]}         & 18.2 & 0.2 \\
       & rec31        & 3415.5                       & (62.1)                                                                          & 3419.3    & (37.0)    & {[}-52.6,& 45.0{]}         & 23.1 & 0.5 \\
\bottomrule
\end{tabular}
%}
\end{table*}
